---
title: "Intro to Statistical Power for Randomized Trials"
subtitle: "Deck 5: Multisite randomized trials"
author: "Eric Hedberg, PhD"
output: 
  ioslides_presentation: 
    fig_retina: null
    highlight: pygments
    incremental: yes
    smart: no
    transition: faster
    widescreen: yes
runtime: shiny
---

```{r, include=FALSE}
source("functions.R")
```

## Many little experiments

In contrast to randomizing clusters, another choice with two-level designs is to randomize units within clusters. 

This essentially means that each cluster, or "site," becomes its own small trial. 

This allows researchers to estimate how treatment effects vary across sites. 

So-called multisite randomized trials are also popular in social and health experiments.

The sample breakdown is this: we have $j = \{1,2\}$ treatments, each with $i = \{1,2,\dots n\}$ units assigned to them within each cluster, and each set of units is assigned to $k = \{1,2,\dots m\}$ clusters (sites). 

## Multisite randomized trials (mixed notation)

As with cluster randomized trials, we can estimate a mixed model to perform the analysis to test the treatment effect. 

The regression model in mixed notation is as follows. The outcome for the $i^{th}$ unit assigned to treatment group $j$ in cluster $k$, $y_{ijk}$, is modeled as a function of a treatment indicator ($T$), a random effect for the cluster $b_{k}$, a random effect of the treatment effect $\left(Tb\right)_{jk}$, and a within-group residual, $e_{ijk}$
$$	y_{ijk} = \gamma_0 + \gamma_1 T_{ijk} + b_{k} + \left(Tb\right)_{jk}+e_{ijk} $$

## Multisite randomized trials (HLM notation)

Exactly how the interaction works and what it means can be better understood with HLM notation. Using HLM notation, we can posit a "level-1" model where the outcome is a function of a cluster-specific control group mean, $\pi_{0k}$, a cluster-specific treatment effect, $\pi_{1k}$, and a within-cluster residual $e_{ijk}$
$$	y_{ijk} = \pi_{0k} + \pi_{1k} T_{ijk}+e_{ijk} $$

## HLM notation for multisite randomized trials (I)

Exactly how the interaction works and what it means can be better understood with HLM notation. Using HLM notation, we can posit a "level-1" model where the outcome is a function of a cluster-specific control group mean, $\pi_{0k}$, a cluster-specific treatment effect, $\pi_{1k}$, and a within-cluster residual $e_{ijk}$

$$	y_{ijk} = \pi_{0k} + \pi_{1k} T_{ijk}+e_{ijk} $$

## HLM notation for multisite randomized trials (II)

Next, we have two "level-2" models. The cluster-specific control group mean is a function of the average of cluster control group means plus a cluster-specific random effect 

$$\pi_{0k} = \gamma_{00} + b_{0k},$$

and similarly, the cluster specific treatment effect is a function of an average of the cluster-specific treatment effects 
$$	\pi_{1k} = \gamma_{10} + b_{1k}$$

## From HLM to mixed

We can put it all together to get to the mixed model, if 

* $y_{ijk} = \pi_{0k} + \pi_{1k} + T_{ijk} + e_{ijk}$, 
* $\pi_{0k} = \gamma_{00} + b_{0k}$, and 
* $\pi_{1k} = \gamma_{10} + b_{1k}$, then

$$y_{ijk} = \gamma_{00} + b_{0k} + \left(\gamma_{10} + b_{1k}\right) T_{ijk}+e_{ijk}$$

so

$$ y_{ijk} = \gamma_{00} + \gamma_{10}T_{ijk} + b_{0k} + b_{1k}T_{ijk}+e_{ijk} $$

similar to the mixed model

$$	y_{ijk} = \gamma_0 + \gamma_1 T_{ijk} + b_{k} + \left(Tb\right)_{jk}+e_{ijk} $$


## Sources of variance

Again, the variance of the random effects are the estimated variance components. 

* The random effect $e_{ik(j)}$ is distributed normal with a mean of zero and standard deviation $\sigma_e$, 
* the random effect $b_{k}$ is distributed normal with a mean of zero and standard deviation $\sigma_b$, and 
* the random effect $\left(Tb\right)_{jk}$ is distributed normal with a mean of zero and standard deviation $\sigma_{\tau b}$.

$$	y_{ijk} = \gamma_0 + \gamma_1 T_{ijk} + b_{k} + \left(Tb\right)_{jk}+e_{ijk} $$

## The t-test for multisite randomized trials (no covariates)

Here is the test, (note that we don't even use the cluster intercept variance)

$$ t = \left(\mu_1-\mu_0\right)\sqrt{\frac{nm}{2\left(\sigma_e^2+\frac{n}{2}\sigma_{\tau b}^2\right)}} $$

However, we do use the cluster level variance to create the scale free parameters. 

## With scale-free parameters

We first work with the difference to create an effect size that employs the total variance, $\delta_b = \frac{\left(\mu_1-\mu_0\right)}{\sigma_b}=\frac{\delta}{\sqrt{\rho_{intra}}}$. 

Next we standardize the treatment variance using the between-cluster mean variance $\upsilon=\frac{\sigma_{\tau b}^2}{\sigma_b^2}$

With these, get the _expected_ test with $df = m - 1$

$$\lambda = \underbrace{\delta}_\text{Effect size}\underbrace{\sqrt{\frac{nm}{2}}}_\text{Sample size}\underbrace{\sqrt{\frac{1}{1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}}}}_\text{Design effect}$$

## The treatment variance ratio's effect on power

```{r, echo = FALSE}
inputPanel(
  sliderInput("n.1", label = "Number units per group per cluster (n)",
              min = 2, max = 50, step = 1, value = 25),
  sliderInput("m.1", label = "Number clusters (m)",
              min = 2, max = 20, step = 1, value = 10),
  sliderInput("es.1", label = "Effect size (delta)",
              min = 0.1, max = 1, value = .3, step = 0.1),
  sliderInput("icc.1", label = "Intraclass correlation",
              min = 0, max = .5, value = .2, step = 0.05),
  sliderInput("upsilon.1", label = "Ratio of v(treat) to v(mean)",
              min = 0, max = 2, value = .2, step = 0.05)
  )

renderPlot({
  exactPowerPlot(type = "msrt", es = input$es.1, 
                 n = input$n.1, 
                 m = input$m.1,
                 icc = input$icc.1,
                 upsilon = input$upsilon.1,
                 marklambda = TRUE)
}, width = 1000, height = 300)
```

## Power analysis for multisite randomized trials (no covariates)

By now, we should be getting the hang of this. Compute the design effect, $1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}$

The test is ($df = m - 1$)

$$ \lambda = \delta\sqrt{\frac{nm/2}{1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}}} $$

## Power analysis for multisite randomized trials (no covariates) example I

Suppose we expected an effect size of $\delta = 0.3$, with $m = 5$ clusters (schools), each with $n = 20$ units (students) in each group. We also expect to have an intraclass correlation of $rho_{intra} = 0.2$ and a ratio of treatment variance to the mean variance of $\upsilon = .5$. We start this code like the other examples

```{r}
m <- 5
n <- 20
df <- m - 1
alpha <- .05
delta <- .3
rho.intra <- .2
upsilon <- .5
D <- 1 + (n/2*upsilon-1)*rho.intra
```

To organize our code, we will code a parameter `D` which is the design effect. 

## Power analysis for multisite randomized trials (no covariates) example II

We then finish our code much the same way as the other example, calculating $\lambda$ and finding the areas under the curve

```{r}
lambda <- delta*sqrt(n*m/2)*sqrt(1/D)
beta <- pt(qt(1-alpha/2,df),df,lambda) - pt(qt(alpha/2,df),df,lambda)
power <- 1-beta
power
```

## Power analysis for multisite randomized trials (no covariates) example III

Here is the complete program

```{r}
m <- 5
n <- 20
df <- m - 1
alpha <- .05
delta <- .3
rho.intra <- .2
upsilon <- .5
D <- 1 + (n/2*upsilon-1)*rho.intra
lambda <- delta*sqrt(n*m/2)*sqrt(1/D)
beta <- pt(qt(1-alpha/2,df),df,lambda) - pt(qt(alpha/2,df),df,lambda)
power <- 1-beta
power
```

## Sample size for multisite randomized trials (no covariates) I

Holding all else constant, adding units past 20 has diminishing returns. This is because the design effect $1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}$ is proportional to the within-cluster sample size. 

As a result, we will focus on finding the number of clusters. 

Like before, we need to employ a normal approximation first, then use that result for a t-distribution based estimate. 

Like our code, we will organize the formulas by noting 

$$D = 1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}$$

## Sample size for multisite randomized trials (no covariates) II

The normal approximation is 

$$ m_z \approx \frac{2\left(z_{critical}-z_{\beta}\right)^2D}{n\delta^2} $$

and then we use the $t$ distribution

$$ m \approx \frac{2\left(t_{(m_z-1)critical}-t_{(m_z-1)\beta}\right)^2D}{n\delta^2} $$

## Sample size for multisite randomized trials (no covariates) example I

Suppose we expected an effect size of $\delta = 0.3$, with $n = 20$ units (students) in each group per cluster. We also expect to have an intraclass correlation of $\rho_{intra} = 0.2$ and a ratio of treatment variance to the mean variance of $\upsilon = .5$. How many clusters do we need to achieve power of 0.8? We start this code like the other examples,

``` {r}
n <- 20
alpha <- .05
power <- .8
delta <- .3
rho.intra <- .2
upsilon <- .5
```

## Sample size for multisite randomized trials (no covariates) example II

We then calculate `D` and find the number of clusters

``` {r}
D <- 1 + (n/2*upsilon-1)*rho.intra
z.crit <- qnorm(1-alpha/2)
z.beta <- qnorm(1-power)
m.z <- 2*(z.crit-z.beta)^2*D/(n*delta^2)
df <- ceiling(m.z) - 1
t.crit <- qt(1-alpha/2, df)
t.beta <- qt(1-power, df)
m <- 2*(t.crit-t.beta)^2*D/(n*delta^2)
m
```

## Sample size for multisite randomized trials (no covariates) example III

Full program (using ";" for line breaks to make it fit)

``` {r}
n <- 20
alpha <- .05; power <- .8
delta <- .3; rho.intra <- .2; upsilon <- .5
D <- 1 + (n/2*upsilon-1)*rho.intra
z.crit <- qnorm(1-alpha/2); z.beta <- qnorm(1-power)
m.z <- 2*(z.crit-z.beta)^2*D/(n*delta^2)
df <- ceiling(m.z) - 1
t.crit <- qt(1-alpha/2, df); t.beta <- qt(1-power, df)
m <- 2*(t.crit-t.beta)^2*D/(n*delta^2)
m
```

## Minimum Detectable Effect Size for multisite randomized trials (no covariates) I

We can again rearrange $\lambda$ for the effect size

$$ \delta_m \approx \sqrt{\frac{2D}{nm}}\left(t_{(m-1)critical}-t_{(m-1)\beta}\right) $$ 
where

$$D = 1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}$$

and since we are already planning on a sample size, we know the $df = m-1$, so no further approximation is necessary

## MDES example for multisite randomized trials (no covariates) I

Suppose we knew that we had $n = 20$ cases per group per cluster and $m = 5$ clusters with an expected intraclass correlation of $\rho_{intra}=0.1$ and a ratio of treatment variance to the mean variance of $\upsilon = .5$, and we wanted the minimum effect size for power of 0.8 ($\beta = 0.2$), for a two-tailed test with $\alpha = 0.05$ 

```{r }
n <- 20; m <- 5; df <- m-1
alpha <- .05
power <- .8
rho.intra <- .1
D <- 1 + (n/2*upsilon-1)*rho.intra
t.crit <- qt(1-alpha/2, df)
t.beta <- qt(1-power, df)
delta.m <- sqrt((2*D)/(m*n))*(t.crit-t.beta)
delta.m
```

## Multisite randomized trials (with coviarates)

As with cluster randomized trials, we can enter covariates at the unit and cluster levels to reduce the variance components and increase power.  

The regression model in mixed notation is as follows. The outcome for the $i^{th}$ unit assigned to treatment group $j$ in cluster $k$, $y_{ijk}$, is modeled as a function of a treatment indicator ($T$), the covariate $x$ and its cluster mean, a random effect for the cluster $b_{k}$, a random effect of the treatment effect $\left(Tb\right)_{jk}$, and a within-group residual, $e_{ijk}$
$$	y_{ijk} = \gamma_0 + \gamma_1 T_{ijk} +\gamma_2\left(x_{ijk}-\bar{x}_k\right)+\gamma_3\bar{x}_k + b_{k}^* + \left(Tb\right)_{jk}^*+e_{ijk}^* $$

This again creates $R^2$ parameters For example, the variance of $e^*_{ik(j)}$ would be $\sigma^{2*}_e=\sigma^2_e\left(1-R^2_{unit}\right)$, and the variance of $\left(Tb\right)_{jk}^*$ would be $\sigma^{2*}_{\tau b} = \sigma^{2}_{\tau b}\left(1-R^2_{treat}\right)$

## The test for multisite randomized trials (with coviarates)

The test with covariates, which has $df = m-1-1$, is

$$ \lambda = \delta\sqrt{\frac{nm/2}{1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}-\left(R_{unit}^2+\left(\frac{n}{2}\upsilon R_{treat}^2-R_{unit}^2\right)\rho_{intra}\right)}} $$

where the design effect is 

$$D = 1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}-\left(R_{unit}^2+\left(\frac{n}{2}\upsilon R_{treat}^2-R_{unit}^2\right)\rho_{intra}\right)$$

## Power analysis for multisite randomized trials (with covariates) example I

Suppose we expected an effect size of $\delta = 0.3$, with $m = 5$ clusters (schools), each with $n = 20$ units (students) in each group. We also expect to have an intraclass correlation of $\rho_{intra} = 0.2$ and a ratio of treatment variance to the mean variance of $\upsilon = .5$. Also, we expect to have a covariate that explains a quarter of the variation at the unit level ($R^2_{unit} = .25$) and half of the treatment variance ($R^2_{treat} = .5$).  We start this code like the other examples

```{r}
m <- 5; n <- 20
df <- m - 1 - 1
alpha <- .05
delta <- .3
rho.intra <- .2; R2.unit <- .25
upsilon <- .5; R2.treat <- .5
D <- 1 + (n/2*upsilon-1)*rho.intra- (R2.unit + (n/2*upsilon*R2.treat-R2.unit)*rho.intra)
```

To organize our code, we will code a parameter `D` which is the design effect. 

## Power analysis for multisite randomized trials (with covariates) example II

We then finish our code much the same way as the other example, calculating $\lambda$ and finding the areas under the curve

```{r}
lambda <- delta*sqrt(n*m/2)*sqrt(1/D)
beta <- pt(qt(1-alpha/2,df),df,lambda) - pt(qt(alpha/2,df),df,lambda)
power <- 1-beta
power
```

## Power analysis for multisite randomized trials (with covariates) example III

Here is the complete program

```{r}
m <- 5; n <- 20; df <- m - 1 - 1
alpha <- .05; delta <- .3
rho.intra <- .2; R2.unit <- .25
upsilon <- .5; R2.treat <- .5
D <- 1 + (n/2*upsilon-1)*rho.intra- (R2.unit + (n/2*upsilon*R2.treat-R2.unit)*rho.intra)
lambda <- delta*sqrt(n*m/2)*sqrt(1/D)
beta <- pt(qt(1-alpha/2,df),df,lambda) - pt(qt(alpha/2,df),df,lambda)
power <- 1-beta
power
```

## Sample size for multisite randomized trials (with covariates) I

Like before, we need to employ a normal approximation first, then use that result for a t-distribution based estimate. 

Like our code, we will organize the formulas by noting 

$$D = 1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}-\left(R_{unit}^2+\left(\frac{n}{2}\upsilon R_{treat}^2-R_{unit}^2\right)\rho_{intra}\right)$$

## Sample size for multisite randomized trials (with covariates) II

The normal approximation is 

$$ m_z \approx \frac{2\left(z_{critical}-z_{\beta}\right)^2D}{n\delta^2} $$

and then we use the $t$ distribution

$$ m \approx \frac{2\left(t_{(m_z-1-1)critical}-t_{(m_z-1-1)\beta}\right)^2D}{n\delta^2} $$

## Sample size for multisite randomized trials (with covariates) example I

Suppose we expected an effect size of $\delta = 0.3$, with $n = 20$ units (students) in each group per cluster. We also expect to have an intraclass correlation of $\rho_{intra} = 0.2$ and a ratio of treatment variance to the mean variance of $\upsilon = .5$. Also, we expect to have a covariate that explains a quarter of the variation at the unit level ($R^2_{unit} = .25$) and half of the treatment variance ($R^2_{treat} = .5$). How many clusters do we need to achieve power of 0.8? We start this code like the other examples,

``` {r}
n <- 20
alpha <- .05
power <- .8
delta <- .3
rho.intra <- .2; R2.unit <- .25
upsilon <- .5; R2.treat <- .5
```

## Sample size for multisite randomized trials (with covariates) example II

We then calculate `D` and find the number of clusters

``` {r}
D <- 1 + (n/2*upsilon-1)*rho.intra- (R2.unit + (n/2*upsilon*R2.treat-R2.unit)*rho.intra)
z.crit <- qnorm(1-alpha/2); z.beta <- qnorm(1-power)
m.z <- 2*(z.crit-z.beta)^2*D/(n*delta^2)
df <- ceiling(m.z) - 1 - 1
t.crit <- qt(1-alpha/2, df); t.beta <- qt(1-power, df)
m <- 2*(t.crit-t.beta)^2*D/(n*delta^2)
m
```

## Sample size for multisite randomized trials (with covariates) example III

Full program (using ";" for line breaks to make it fit)

``` {r}
n <- 20
alpha <- .05; power <- .8; delta <- .3
rho.intra <- .2; R2.unit <- .25; upsilon <- .5; R2.treat <- .5
D <- 1 + (n/2*upsilon-1)*rho.intra- (R2.unit + (n/2*upsilon*R2.treat-R2.unit)*rho.intra)
z.crit <- qnorm(1-alpha/2); z.beta <- qnorm(1-power)
m.z <- 2*(z.crit-z.beta)^2*D/(n*delta^2)
df <- ceiling(m.z) - 1 - 1
t.crit <- qt(1-alpha/2, df); t.beta <- qt(1-power, df)
m <- 2*(t.crit-t.beta)^2*D/(n*delta^2)
m
```

## Minimum detectable effect size for multisite randomized trials (with covariates) I

We can again rearrange $\lambda$ for the effect size

$$ \delta_m \approx \sqrt{\frac{2D}{nm}}\left(t_{(m-1-1)critical}-t_{(m-1-1)\beta}\right) $$ 
where

$$D = 1+\left(\frac{n}{2}\upsilon-1\right)\rho_{intra}-\left(R_{unit}^2+\left(\frac{n}{2}\upsilon R_{treat}^2-R_{unit}^2\right)\rho_{intra}\right)$$

and since we are already planning on a sample size, we know the $df = m-1-1$, so no further approximation is necessary

## MDES example for multisite randomized trials (with covariates) I

Suppose we knew that we had $n = 20$ cases per group per cluster and $m = 5$ clusters with an expected intraclass correlation of $\rho_{intra}=0.1$ and a ratio of treatment variance to the mean variance of $\upsilon = .5$. Also, we expect to have a covariate that explains a quarter of the variation at the unit level ($R^2_{unit} = .25$) and half of the treatment variance ($R^2_{treat} = .5$). We want the minimum effect size for power of 0.8 ($\beta = 0.2$), for a two-tailed test with $\alpha = 0.05$ 

```{r }
n <- 20; m <- 5; df <- m-1-1
alpha <- .05; power <- .8; rho.intra <- .1
D <- 1 + (n/2*upsilon-1)*rho.intra- (R2.unit + (n/2*upsilon*R2.treat-R2.unit)*rho.intra)
t.crit <- qt(1-alpha/2, df); t.beta <- qt(1-power, df)
delta.m <- sqrt((2*D)/(m*n))*(t.crit-t.beta)
delta.m
```

## Explore!

```{r, echo = FALSE}
inputPanel(
  sliderInput("n.2", label = "Number units per group per cluster (n)",
              min = 2, max = 50, step = 1, value = 25),
  sliderInput("m.2", label = "Number clusters (m)",
              min = 2, max = 20, step = 1, value = 10),
  sliderInput("es.2", label = "Effect size (delta)",
              min = 0.1, max = 1, value = .3, step = 0.1),
  sliderInput("icc.2", label = "Intraclass correlation",
              min = 0, max = .5, value = .2, step = 0.05),
  sliderInput("upsilon.2", label = "Ratio of v(treat) to v(mean)",
              min = 0, max = 2, value = .2, step = 0.05),
  sliderInput("r2u.2", label = "R-square unit",
              min = 0, max = 1, value = .2, step = 0.05),
  sliderInput("r2t.2", label = "R-square treatment",
              min = 0, max = 1, value = .2, step = 0.05)
  )

renderPlot({
  exactPowerPlot(type = "msrt", 
                 es = input$es.2, 
                 n = input$n.2, 
                 m = input$m.2,
                 icc = input$icc.2,
                 upsilon = input$upsilon.2,
                 cov = "Yes",
                 r2u = input$r2u.2,
                 r2t = input$r2t.2,
                 marklambda = TRUE)
}, width = 1000, height = 300)
```

